Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

### 2024-01-30 manuscript/manuscript.qmd
Revisions and Rebuttal
Revisions to manuscript + Rebuttal draft

Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008]. However, the substantial increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all. This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.

Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification [@bianco_machine_2019, @mutanu_review_2022, @stowell_computational_2022]. In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including marine mammals [@frainer_automatic_2023, @frasier_automated_2017, @li_model-based_2020, @roman_ruiz_fin_2023, @zhong_beluga_2020]. Neural networks typically require significant sample sizes for all classes, although there are increasingly new tools to enhance data or combined disparate datasets [@dufourq_passive_2022, @nanni_data_2020, @nolasco_learning_2023, @padovese_adapting_2023, @shorten_survey_2019]. There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset, although development of more user friendly tools are in development [@kahl_birdnet_2021, @kirsebom_meridian_2022]. Despite the great potential for deep learning methods, modification of existing machine learning methods with a lower barrier to entry may be sufficient for many bioacoustic classification tasks.

Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification [@lopez-baucells_stronger_2019, @ross_random_2014]. BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a hierarchical Random Forest [@breiman_random_2001] event classifier originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017]. BANTER combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter. The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors used to identify calls from any species. BANTER is easy to use and is readily available as a stand-alone R package [@archer_banter_2022].

@@

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018]. However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013]. Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020]. Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.

@@

Experienced acoustic analysts used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale. Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@r_core_team_r_2022].

Standard click metrics describing the spectral and temporal characteristics were measured from echolocation pulses using default values in *PAMpal* [@sakai_pampal_2023]. Inter-pulse interval (IPI) was calculated using the 'calculateICI' function in *PAMpal* and the mode of the IPI was identified for each beaked whale acoustic event. In addition, for the EPacific data, we used the 'matchEnvData' function in *PAMpal* to include ERDAPP environmental data [@simons_erddap_2022], including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41). A full consideration of environmental data for all regions was beyond the scope of this study.

@@

BANTER models were created using the *banter* package [@archer_banter_2022] in R v2022.07.02 [@r_core_team_r_2022]. Output from PAMGuard click detector modules were used to create the BANTER detector models, for a total of 6 detector models (one for each click detector used). For each detector model, BANTER creates a mean of the species classification probabilities for each call type, which is applied to the BANTER event model (in conjunction with any event-level variables). A minimum of two events per species and the complete suite of variables (no missing data) were required for training and testing of the classification model. Event level variables included IPI and environmental variables (for the EPacific dataset). The Random Forest models that BANTER is based on are largely driven by two parameters, the number of trees in the forest (*ntree*) and the number of samples randomly drawn to build each tree (*sampsize*). Small values of *sampsize* were selected to improve computational performance, but large values of *ntree* were used to obtain models with stable classification results [@rankin_banter_2021]. Model results were summarized using the *rfpermute* package [@archer_rfpermute_2022].

@@

The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset. All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%. Other studies on machine learning acoustic classification of beaked whales have yielded stronger classification results, although these studies included fewer beaked whale species classes and much larger sample sizes for each species class \[@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022\]. While there is great room for improvement in our models, and we suspect that modest increases in sample sizes for some species will result in significant improvements to the model results, it is likely that the low sample sizes presented in these datasets would be insufficient to develop and test deep learning models.

@@

In addition to poor classification of Gervais' in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais', and the majority of the votes were for True's beaked whales. Analysts' notes indicated uncertainty in manual species classification of this event in this dataset. While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influential if there are inaccuracies in the training data.

These two examples of small sample sizes highlight a conflict in preferred protocol. Ideally, (1) all [species]{.underline} would be included in a classification model, so as to better represent the local species diversity, (2) all [events]{.underline} would ideally be included in the classification model, so as to better represent the variability found in the area, and (3) only confident 'ground truth' classifications would be considered in the training model. Unfortunately, in species where identity must be determined based on call characteristics (rather than visual confirmation of species identity in the field), it can sometimes be difficult to confidently determine species identity when calls are highly variable and do not include at least one call that provides high confidence of species identity. In the case of beaked whales, we recommend that training data include a high level of confidence for inclusion. This is especially critical for species with small sample sizes. An alternative is to require agreement from multiple analysts. This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified. For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysts did not identify these in the EPacific data presented here).

@@

The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies. Thanks to Liam Mueller-Brennan for their help with data analysis. Sea surface temperature data were provided by JPL under support by the NASA MEaSUREs program. Funding for data collection and analysis were provided by the U.S. Navy, the Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration. The AMAPPS study (NAtlantic/SAtantic) was funded in part by the U.S. Department of the Interior, Bureau of Ocean Energy Management, Environmental Studies Program, Washington, DC, through Inter-Agency Agreement Number M10PG00075 with the National Marine Fisheries Service as the Atlantic Marine Assessment Program for Protected Species. The HICEAS 2017 Survey (Hawaii) was funded by Bureau of Ocean Energy Management under (IAA M17PG00024) and NMFS Pacific Islands Fisheries Science Center. The CCES Survey (EPacific) was funded by Bureau of Ocean Energy Management (No. M17PG00025), U.S. Navy Pacific Fleet Environmental Readiness Division (IAA No. N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center. The manuscript was improved thanks to reviews from Cory Hom-Weaver.

# Figures and tables

```{r, natlantic}
#| label: fig-natlantic
#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation pulses and inter-pulse interval (EC_IPI). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 100%
#| echo: false
confuse <- magick::image_read('manuscript_files/natlantic_ici_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "50x130")%>%
  image_annotate("a) Confusion Matrix", size=100, color = "black")
vote <- magick::image_read('manuscript_files/natlantic_ici_votes.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("d) Vote Plot", size=100, color = "black")
prox <- magick::image_read('manuscript_files/natlantic_ici_proximity.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_annotate("b) Proximity Plot", size=100, color = "black")
heat <- magick::image_read('manuscript_files/natlantic_ici_importance.png')%>%
  image_border(color="#ffffff", geometry = "270x130")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=100, color = "black")

>>>>

Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013; @bittle_review_2013; @rankin_source_2005; @soldevilla_classification_2008]. However, the substantial increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all. This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.

Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification [@bianco_machine_2019; @mutanu_review_2022; @stowell_computational_2022]. In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including marine mammals [@frainer_automatic_2023; @frasier_automated_2017; @li_model-based_2020; @roman_ruiz_fin_2023; @zhong_beluga_2020]. Neural networks typically require significant sample sizes for all classes, although there are increasingly new tools to enhance data or combined disparate datasets [@dufourq_passive_2022; @nanni_data_2020; @nolasco_learning_2023; @padovese_adapting_2023; @shorten_survey_2019]. There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset, although development of more user friendly tools are in development [@kahl_birdnet_2021; @kirsebom_meridian_2022]. Despite the great potential for deep learning methods, modification of existing machine learning methods with a lower barrier to entry may be sufficient for many bioacoustic classification tasks.

Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification [@lopez-baucells_stronger_2019; @ross_random_2014]. BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a hierarchical Random Forest [@breiman_random_2001] event classifier originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017]. BANTER combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter. The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors used to identify calls from any species. BANTER is easy to use and is readily available as a stand-alone R package [@archer_banter_2022].

@@

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018]. However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013]. Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022; @baumann-pickering_spatio-temporal_2014; @simonis_passive_2020]. Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.

@@

Experienced acoustic analysts used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale. Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@team_r_2013].

Standard click metrics describing the spectral and temporal characteristics were measured from echolocation pulses using default values in *PAMpal* [@sakai_pampal_2021]. Inter-pulse interval (IPI) was calculated using the 'calculateICI' function in *PAMpal* and the mode of the IPI was identified for each beaked whale acoustic event. In addition, for the EPacific data, we used the 'matchEnvData' function in *PAMpal* to include ERDAPP environmental data [@simons_erddap_2022], including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41). A full consideration of environmental data for all regions was beyond the scope of this study.

@@

BANTER models were created using the *banter* package [@archer_banter_2022] in R v2022.07.02 [@team_r_2013]. Output from PAMGuard click detector modules were used to create the BANTER detector models, for a total of 6 detector models (one for each click detector used). For each detector model, BANTER creates a mean of the species classification probabilities for each call type, which is applied to the BANTER event model (in conjunction with any event-level variables). A minimum of two events per species and the complete suite of variables (no missing data) were required for training and testing of the classification model. Event level variables included IPI and environmental variables (for the EPacific dataset). The Random Forest models that BANTER is based on are largely driven by two parameters, the number of trees in the forest (*ntree*) and the number of samples randomly drawn to build each tree (*sampsize*). Small values of *sampsize* were selected to improve computational performance, but large values of *ntree* were used to obtain models with stable classification results [@rankin_banter_2021]. Model results were summarized using the *rfpermute* package [@archer_rfpermute_2022].

@@

Development of automated acoustic classifiers is critical for efficient analysis of large passive acoustic datasets; these classification routines require representative training data created by experienced acoustic analysts. While beaked whales species can be identified based on their stereotyped frequency-modulated echolocation clicks [@baumann-pickering_species-specific_2013], not all acoustic events include sounds that can be readily attributed to a specific species. Moreover, there are a number of beaked whales where little or nothing is known about their acoustic repertoire and new species have been detected in recent years [@deangelis_description_2018; @barlow_unique_2022]. Therefore, development of acoustic classifiers for beaked whales should accommodate small and/or unbalanced sample sizes and have sufficient flexibility to allow for variable call characteristics.

The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset. All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%. Recent beaked whale classification studies using deep learning methods exhibit more robust results [@li_model-based_2020-1; @lebien_species-level_2018; @cohen_identification_2022; @ziegenhorn_discriminating_2022]; however, these models may not be appropriate for our dataset. Several studies included incomplete beaked whale species for the study area (@li_model-based_2020-1, @ziegenhorn_discriminating_2022) and all studies has sample large sample sizes (\>200) for each species class. The deep learning model developed by @cohen_identification_2022 included strong classification results with all beaked whale species found in our Atlantic models; however, it is unclear how this dataset trained on seafloor recorders would perform with towed array datasets. The BANTER models presented here only require a minimum of two samples per class and can therefore include all species within each region, making them useful to assess the presence of all beaked whales in the region. While deep learning methods may provide improved results once there are sufficient sample sizes for all relevant species within a study area, our results suggest that reasonable classification results can be obtained for all species using BANTER classification methods with as few as two samples per class.

@@

In addition to poor classification of Gervais' in the NAtlantic study area, the plot votes show that for one of the four events, only a few of the 100,000 trees 'voted' for Gervais', and the majority of the votes were for True's beaked whales. Analysts' notes indicated uncertainty in manual species classification of this event in this dataset. While small sample size can at times provide good classification scores, as we found for Sowerby's beaked whale in this SAtlantic dataset, they can be heavily influenced by inaccuracies in the training data.

These two examples of small sample sizes highlight a conflict in preferred protocol. Ideally, (1) all [species]{.underline} would be included in a classification model, so as to better represent the local species diversity, (2) all [events]{.underline} would ideally be included in the classification model, so as to better represent the variability found in the area, and (3) only confident 'ground truth' classifications would be considered in the training model. Unfortunately, in species where identity must be determined based on call characteristics (rather than visual confirmation of species identity in the field), it can sometimes be difficult to confidently determine species identity when calls are highly variable and do not include at least one call that provides high confidence of species identity. In the case of beaked whales, we recommend that training data include a high level of confidence for inclusion. This is especially critical for species with small sample sizes. An alternative is to require agreement from multiple analysts. This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified. For example, @barlow_unique_2022 recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysts did not identify these in the EPacific data presented here).

@@

The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies. Thanks to Liam Mueller-Brennan for their help with data analysis. Sea surface temperature data were provided by JPL under support by the NASA MEaSUREs program. Funding for data collection and analysis were provided by the U.S. Navy, the Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration. The AMAPPS study (NAtlantic/SAtantic) was funded in part by the U.S. Department of the Interior, Bureau of Ocean Energy Management, Environmental Studies Program, Washington, DC, through Inter-Agency Agreement Number M10PG00075 with the National Marine Fisheries Service as the Atlantic Marine Assessment Program for Protected Species. The HICEAS 2017 Survey (Hawaii) was funded by Bureau of Ocean Energy Management under (IAA M17PG00024) and NMFS Pacific Islands Fisheries Science Center. The CCES Survey (EPacific) was funded by Bureau of Ocean Energy Management (No. M17PG00025), U.S. Navy Pacific Fleet Environmental Readiness Division (IAA No. N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center. The manuscript was improved thanks to reviews from Cory Hom-Weaver and an anonymous reviewer.

{{< pagebreak >}}
# Figures and tables

```{r, natlantic}
#| label: fig-natlantic
#| fig-cap: BANTER classification results from the NAtlantic dataset including echolocation pulses and inter-pulse interval (EC_IPI). Confusion matrix (a) provides the percent correct classification for each species (pct.correct), lower confidence intervals (LCI_0.95), upper confidence intervals (UCI_0.95), and priors (expected error rate). Proximity plot (b) for species events from BANTER model (central dot color represents true species identity; color of circle surrounding dot represents BANTER species classification). Heat map (c) for ranks of ten most important variables; colors scale from most important predictors (dark red) to least important predictors (dark blue). Vote Plot (d) shows the vote distribution for each event (vertical slice) for each species; distribution of votes by species is shown by their representative color.
#| fig-width: 10
#| fig-height: 10
#| fig-align: center
#| out-width: 200%
#| echo: false
confuse <- magick::image_read('manuscript_files/natlantic_ici_confuseMatrix.png')%>%
  image_border(color="#ffffff", geometry = "80x208")%>%
  image_annotate("a) Confusion Matrix", size=120, color = "black")
vote <- magick::image_read('manuscript_files/natlantic_ici_votes.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("d) Vote Plot", size=120, color = "black")
prox <- magick::image_read('manuscript_files/natlantic_ici_proximity.png')%>%
  image_border(color="#ffffff", geometry = "260x125")%>%
  image_annotate("b) Proximity Plot", size=120, color = "black")
heat <- magick::image_read('manuscript_files/natlantic_ici_importance.png')%>%
  image_border(color="#ffffff", geometry = "110x260")%>%
  image_scale("3300")%>%
  image_annotate("c) Importance Heat Map", size=120, color = "black")
